### 2024-07-17 parts/prelim.tex
update citations

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a function.
    The following function 
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is \kl{$\Rel$-polyregular}:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is furthermore \kl{$\Nat$-polyregular}.
    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
    If the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \kl{star-free $\Rel$-polyregular}
    (resp. \kl{star-free $\Nat$-polyregular}).
\end{definition}

In order to illustrate the above definition, let us provide an example of a
\kl{$\Nat$-polyregular function} computed using a finite monoid.

\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
\end{example}
\begin{proof}
    Let us define $M \defined (\set{1}, \times)$ which is 
    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
    defined by $\mu(w) \defined 1$, and
    $\pi \colon M^2 \to \Nat$
    that is the constant function equal to $1$.
    We check that for all $w \in \Sigma^*$:
    \begin{equation*}
        \pi^\dagger(w)
        =
        \sum_{uv = w} 1
        =
        \card{w} + 1
        = f(w)
        \quad . 
        \qedhere
    \end{equation*}
\end{proof}

\AP One of the appeals of $\NPoly$ and $\ZPoly$ are the numerous
characterizations of these classes in terms of logic, weighted automata, and
the larger class of \kl{polyregular functions} \cite{CDTL23,DOUE23}. In this
paper, the main focus will be the connection to weighted automata, which is
based on the notion of \emph{growth rate}. The \intro{growth rate} of a
function $f \colon \Sigma^* \to \Rel$ is defined as the minimal $k$ such that
$\card{f(w)} = \bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that
the function $f$ has \intro{polynomial growth}. It turns out that for all $k
\in \Nat$, $\ZPoly[k]$ (resp. $\NPoly[k]$) are precisely functions in $\ZRat$
(resp. in $\NRat)$ that have growth rate at most $k$ \cite{CDTL23}.

\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
define the classes of \intro{$\Rel $-polyregular functions} $\intro*\ZPoly[k]$,
\intro{$\Nat $-polyregular functions} $\intro*\NPoly[k]$, \intro{star-free
$\Rel $-polyregular functions} $\intro*\ZSF[k]$, and \intro{star-free $\Nat
$-polyregular functions} $\intro*\NSF[k]$, as follows \cite[Theorems 5.15 and
7.10]{DOUE23}: in the base case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  =
\reintro*\ZSF[-1] = \reintro*\ZPoly[-1]$ is the singleton containing the
constant function equal to $0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the
set of functions \kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
where the transducer is furthermore required to be \kl{counter-free}. Then,
$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined analogously.
In the rest of the paper, we will always assume that
\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
have an $\Nat$-valued output functions, making a clean separation between
$\NPoly$ and $\ZPoly$. that \kl{represents} $P$.

Because $f$ has \kl{polynomial
growth}, it is in fact a \kl{$\Nat$-polyregular function} (\cite[Theorem
5.22]{DOUE23}).
\begin{lemma}[CITE]
    \label{polyregular-polynomial-growth:lemma}
    Let $f \in \ZRat$. The following are equivalent:
    \begin{enumerate}
        \item $f \in \ZPoly[k]$.
        \item $f$ has polynomial growth of degree at most $k$.
    \end{enumerate}
    And similarly for $\NPoly$ and $\NRat$.
\end{lemma}

Let us state the crucial compositional properties of
\kl{$\Rel$-polyregular functions} that will be used in this paper
to construct \kl{$\Rel$-polyregular functions}.
\begin{lemma}[CITE]
    \label{stability-polyregular:lemma}
    The class $\NPoly[k]$ (resp. $\ZPoly[k]$) is stable under the operations of
    addition, multiplication, Cauchy product,
    and ``value pinning''.
\end{lemma}

@@

\begin{lemma}
    \label{decidable-commutative-poly:lemma}
    \label{decidable-commutative-rat:lemma}
    Let $f \in \ZRat$. One can decide if 
    $f$
    is \kl{commutative}.
\end{lemma}
\begin{proof}
    Remark that the group of permutations of $\set{1, \dots, n}$ is generated by
    the cycle $c \defined (n,1, \dots, n-1)$ and the transposition $t \defined (1, 2)$.
    As a consequence, a function $f$ is commutative if and only if
    $f \circ c = f = f \circ t$.
    When $f$ is a rational series,
    $f \circ c$ and $f \circ t$ are both rational series that can be
    effectively computed from $f$, and since equivalence
    of rational series is decidable 
    \cite[\textbf{WHERE}]{BERE10}
    we have obtained a decision procedure.
\end{proof}

>>>>

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $k \in \Nat$. The set $\intro*\ZPoly[k]$ of polyregular
    \intro{$\Rel$-polyregular functions} of degree at most $k$,
    is the set of functions $f \colon \Sigma^* \to \Rel$
    such that
    there exists a finite monoid $M$,
    a morphism $\mu \colon \Sigma^* \to M$,
    and a function $\pi \colon M^{k+1} \to \Rel$
    satisfying:
    \begin{equation*}
        f \defined
        \sum_{w = u_1 \cdots u_{k+1}} \pi(\mu(u_1), \dots, \mu(u_{k+1}))
        \quad .
    \end{equation*}
    We call $\pi$ the \emph{production function} of $f$.
    If the function $\pi$ has codomain $\Nat$,
    then $f$ is \intro{$\Nat$-polyregular} of degree at most $k$,
    i.e., $f \in \intro*\NPoly[k]$.
    If the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \kl{star-free $\Rel$-polyregular}
    ($\intro*\ZSF[k]$), resp. \kl{star-free $\Nat$-polyregular} ($\intro*\NSF[k]$).
\end{definition}

\AP We complete definition \cref{nat-rel-poly:def} by defining $\intro*\NPoly =
\bigcup_{k \in \Nat} \intro*\NPoly[k]$, and similarly for $\ZPoly$, $\NSF$, and
$\ZSF$. In order to illustrate these definitions, let us provide an example of
a \kl{$\Nat$-polyregular function} computed using a finite monoid.

\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF[1]$.
\end{example}
\begin{proof}
    Let us define $M \defined (\set{1}, \times)$ which is 
    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
    defined by $\mu(w) \defined 1$, and
    $\pi \colon M^2 \to \Nat$
    that is the constant function equal to $1$.
    We check that for all $w \in \Sigma^*$:
    \begin{equation*}
        \pi^\dagger(w)
        =
        \sum_{uv = w} 1
        =
        \card{w} + 1
        = f(w)
        \quad . 
        \qedhere
    \end{equation*}
\end{proof}

\AP One of the appeals of $\NPoly$ and $\ZPoly$ are the numerous
characterizations of these classes in terms of logic, weighted automata, and
the larger class of \kl{polyregular functions} \cite{CDTL23,DOUE23}. In this
paper, the main focus will be the connection to weighted automata, which is
based on the notion of \emph{growth rate}. The \intro{growth rate} of a
function $f \colon \Sigma^* \to \Rel$ is defined as the minimal $k$ such that
$\card{f(w)} = \bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that
the function $f$ has \intro{polynomial growth}. It turns out that for all $k
\in \Nat$, $\ZPoly[k]$ (resp. $\NPoly[k]$) are precisely functions in $\ZRat$
(resp. in $\NRat)$ that have growth rate at most $k$.

\begin{lemma}[{\cite[Theorem 5.22]{DOUE23}}]
    \label{polyregular-polynomial-growth:lemma}
    Let $f \in \ZRat$. The following are equivalent:
    \begin{enumerate}
        \item $f \in \ZPoly[k]$.
        \item $f$ has polynomial growth of degree at most $k$.
    \end{enumerate}
    And similarly for $\NPoly$ and $\NRat$.
\end{lemma}

Let us state the crucial compositional properties of
\kl{$\Rel$-polyregular functions} that will be used in this paper
to construct \kl{$\Rel$-polyregular functions}.
\begin{lemma}[{\cite[Theorem II.20]{CDTL23}}]
    \label{stability-polyregular:lemma}
    The class $\NPoly[k]$ (resp. $\ZPoly[k]$) is stable under the operations of
    addition, multiplication, Cauchy product,
    and ``value pinning''.
\end{lemma}

@@

\begin{lemma}
    \label{decidable-commutative-poly:lemma}
    \label{decidable-commutative-rat:lemma}
    Let $f \in \ZRat$. One can decide if 
    $f$
    is \kl{commutative}.
\end{lemma}
\begin{proof}
    Remark that the group of permutations of $\set{1, \dots, n}$ is generated by
    the cycle $c \defined (n,1, \dots, n-1)$ and the transposition $t \defined (1, 2)$.
    As a consequence, a function $f$ is commutative if and only if
    $f \circ c = f = f \circ t$.
    When $f$ is a rational series,
    $f \circ c$ and $f \circ t$ are both rational series that can be
    effectively computed from $f$, and since equivalence
    of rational series is decidable 
    \cite[Corollary 3.6]{BERE10},
    we have obtained a decision procedure.
\end{proof}
